---
title: "Recreate this document from scratch"
author: "*Your name here*"
format:
  html:
    embed-resources: true
  pdf: default
editor: source
toc: true
---

## Assignment

Create a `.qmd` file that, when rendered, matches this document *exactly*. This will require you to practice the most important Markdown and Quarto features discussed in the reading and in lecture.

## Introduction

Below, we explore the `mtcars` dataset, which can be accessed through the `datasets` package.

You can find the documentation for `mtcars` [here](https://search.r-project.org/R/refmans/datasets/html/mtcars.html).^[Find the list of all of the datasets available in `datasets` [here](https://search.r-project.org/R/refmans/datasets/html/00Index.html).]

### A blockquote about cars

> I live my life a quarter mile at a time.^[Vin Diesel as Dom in *The Fast and the Furious* (2001).]

### Setup

```{r}
#| echo: true
#| warning: false

library(tidyverse)
library(datasets)
```

## About the data

### Sample

Here are the first three rows of `mtcars`:

```{r}
#| echo: false

mtcars |>
  slice_head(n = 3)
```

### Description

- The `mtcars` data was extracted from a 1974 issue of *Motor Trend* magazine.
- It contains **32 rows**, each representing a car model.
  
### Data dictionary

Columns we will analyze include:

| Column Name | Description                | Unit                 |
|-------------|----------------------------|----------------------|
| `mpg`       | Fuel efficiency            | Miles per gallon (US)|
| `wt`        | Weight of car              | Thousands of pounds  |
| `cyl`       | Engine cylinders           | Count                |

### Descriptive statistics

:::{.callout-important}
The values below are calculated **inline**!
:::

```{r}
#| echo: false


mean_mpg <- mtcars |>
  summarize(mean_mpg = mean(mpg)) |>
  round(2) |>
  pull(mean_mpg)

mean_wt <- mtcars |>
  summarize(mean_wt = mean(wt)) |>
  round(2) |>
  pull(mean_wt) * 1000
```

- **Average mpg**: `{r} mean_mpg`
- **Average weight**: `{r} mean_wt`

## Visualizations

### Scatterplot

```{r}
#| label: fig-mtcars-scatter
#| fig-cap: "Scatterplot showing the relationship between miles per gallon and weight in `mtcars`."
#| echo: false

mtcars |>
  ggplot(aes(x = wt, y = mpg)) +
  geom_point() +
  labs(
    title = "Fuel efficiency in relation to weight",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  )
```

### Histogram

@fig-hist uses the line of code below to show the number of cars in each `mpg` bin:

```{r}
#| eval: false
#| echo: true

geom_histogram(bins = 10)
```

```{r}
#| echo: false
#| label: fig-hist
#| fig-cap: "Cars by fuel efficiency (binned)"


mtcars |>
  ggplot(aes(x = mpg)) +
  geom_histogram(bins = 10) +
  labs(
    title = "Histogram of mpg",
    subtitle = "(10 bins)",
    x = "Miles per gallon",
    y = "Frequency"
  )
```

### Boxplot

```{r}
#| label: fig-mtcars-box
#| fig-cap: "Boxplot showing the relationship between the number of cylinders and miles per gallon in `mtcars`."
#| echo: true


mtcars |>
  ggplot(aes(x = as_factor(cyl), y = mpg)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of mpg by number of cylinders",
    x = "Number of cylinders",
    y = "Miles per gallon"
  )
```


### Observations from visualizations

1. @fig-mtcars-scatter suggests that higher weight is associated with *lower* fuel efficiency.
2. @fig-hist suggests that, in 1974, most of the cars that *Motor Trend* covered got between 15 and 25 miles per gallon.^[However, a small number of cars got 30+ miles per gallon, which is impressive for the time.]
3. @fig-mtcars-box suggests that the number of cylinders in a car's engine may also be associated with fuel efficiency.

## Conclusion

`mtcars` illustrates trade-offs between size, power, and efficiency in cars from about 50 years ago.
