---
title: "Regularization"
subtitle: "DSST389: Advanced Data Science"
author: "Erik Fredner"
institute: "University of Richmond"
date: today
bibliography: /Users/erik/book/references.bib
csl: /Users/erik/code/styles/chicago-fullnote-bibliography.csl
execute:
  echo: true
  warning: false
  message: false
format:
  revealjs:
    logo: "images/by-sa.png"
    footer: "https://fredner.org"
    embed-resources: true
    scrollable: true
    toc: true
    toc-depth: 2
    slide-level: 4
    slide-number: true
    preview-links: auto
    mermaid:
      theme: neutral
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: console
---

## Regularization

Elastic net regularization combines the strengths of ridge and lasso regression penalties for linear models. It simultaneously shrinks coefficients (like ridge) and pushes some of them to zero (like lasso).

This balance helps the model handle highly correlated features, avoid overfitting, and maintain interpretability by **filtering out less important variables**.

### Example use cases

1. Predicting the price of a house based on its features (e.g., Zillow)
2. Predicting the likelihood of a customer to churn (e.g., Netflix)
3. Predicting whether an email is spam (e.g., Gmail)

### Intuitive example: Political favorability

| year | zip_code | med_income | avg_wind_speed | dem_fav |
|-----:|---------:|--------------:|---------------:|----------------------:|
| 2020 |   90210  |        120000 |            5.2 |                  0.45 |
| 2020 |   60601  |         85000 |            4.3 |                  0.46 |
| 2021 |   94105  |        110000 |            3.9 |                  0.50 |
| 2021 |   10001  |         95000 |            5.7 |                  0.48 |
| 2022 |   30301  |         70000 |            6.1 |                  0.43 |

#### Intuitive analysis

We would expect average wind speed to be *less* predictive of Democratic favorability than other variables in this table, like year, ZIP code, and income.

## Why regularize?

Linear regression makes several strong assumptions that regularization can help address:

- A linear relationship exists
- The data set has more observations than features
  - i.e., in a tidy data set, more rows than columns
- No multicollinearity exists between features

### What are we doing when we are regularizing?

We are adding a **penalty term**---ridge, lasso, or both, which is the elastic net---to the sum of the squared errors (i.e., the sum of the squared residuals) in our linear model.

The penalty term drives coefficients *down*. Without regularization, coefficients can grow large, leading to overfitting.

## Example: Predicting baseball salaries based on batting statistics

Which aspects of a baseball player's performance at the plate best predict their salary?

### Data: `Hitters`

```{r}
#| echo: false

library(tidyverse)
library(tidymodels)
library(ISLR2)
library(fastDummies)
data(Hitters, package = "ISLR2")
set.seed(123)
theme_set(theme_minimal())

Hitters <- Hitters |>
  drop_na(Salary)
```

```{r}
Hitters |> glimpse()
```

### `Hitters` salary distributions

```{r}
#| echo: false

Hitters |>
  ggplot(aes(Salary)) +
  geom_density(fill = "pink", alpha = 0.2) +
  scale_x_continuous(labels = scales::dollar_format(scale = 1000)) +
  labs(x = "Salary")
```

A small number of hitters have significantly higher salaries than most.

### How well do errors predict salary?

```{r}
#| echo: false

Hitters |>
  ggplot(aes(Errors, Salary)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_segment(
    aes(
      xend = Errors,
      yend = predict(lm(Salary ~ Errors, data = Hitters))
    ),
    alpha = 0.2
  ) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1000)) +
  labs(x = "Errors", y = "Salary (USD, thousands)")
```

### How well does getting walked predict salary?

```{r}
#| echo: false

Hitters |>
  ggplot(aes(CWalks, Salary)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  geom_segment(
    aes(
      xend = CWalks,
      yend = predict(lm(Salary ~ CWalks, data = Hitters))
    ),
    alpha = 0.2
  ) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1000)) +
  labs(x = "Career Walks", y = "Salary (USD, thousands)")
```

### How well do career runs predict salary?

```{r}
#| echo: false

Hitters |>
  ggplot(aes(CRuns, Salary)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "yellow") +
  geom_segment(
    aes(
      xend = CRuns,
      yend = predict(lm(Salary ~ CRuns, data = Hitters))
    ),
    alpha = 0.2
  ) +
  scale_y_continuous(labels = scales::dollar_format(scale = 1000)) +
  labs(x = "Career Runs", y = "Salary (USD, thousands)")
```

### Career Runs, Home Runs, and Salary

- Career runs and salary appear to be positively correlated
- Career runs and home runs are *necessarily* correlated
  - Every home run is at least one additional run

```{r}
Hitters |>
  select(CRuns, CHmRun, Salary) |>
  cor() |> # calculate correlation
  round(2)
```

### Given collinearity, which features should we choose?

```{r}
#| echo: false

Hitters |>
  select(where(is.numeric)) |>
  cor() |>
  corrplot::corrplot()
```

### Deciding on regularization approach

- Ridge is not the best choice here because it does not eliminate any of the features
- Lasso is a good choice because it can eliminate some features entirely
- Elastic net regression (ENET) is another good choice because it can eliminate some features *and* shrink others

## Lasso on `Hitters`

The [`glmnet` package](https://glmnet.stanford.edu/index.html) implements these regularization techniques. We can use it in combination with the `tidymodels` package we used with KNN.

### Split our data

```{r}
hitters_split <- initial_split(Hitters, prop = 0.8, strata = Salary)
hitters_train <- training(hitters_split)
hitters_test <- testing(hitters_split)
```

We are stratifying by a **continuous variable** (salary) to ensure that the distribution of salaries is similar in the training and testing sets.

### Create our lasso model

When using the `linear_reg()` function, we need to set the `mixture` parameter:

- `mixture = 0`: Ridge
- `mixture = 1`: Lasso
- `0 < mixture < 1`: Elastic net

We also need to set a `penalty`, which is the lambda (λ) value that controls the strength of the regularization.

```{r}
lasso_model <- linear_reg(penalty = 0.5, mixture = 1) |>
  set_mode("regression") |>
  set_engine("glmnet")
```

### Create our recipe

```{r}
lasso_recipe <- recipe(Salary ~ ., data = hitters_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())
```

#### `step_dummy(all_nominal_predictors())`?

This converts categorical variables into [dummy variables](https://en.wikipedia.org/wiki/Dummy_variable_(statistics)). `Hitters` has a categorical variable for whether players were in the American or National league:

```{r}
#| echo: false

hitters_train |>
  select(League) |>
  slice_head(n = 4)
```

After conversion, the `League` variables will look like this:

```{r}
#| echo: false

hitters_train |>
  select(League) |>
  slice_head(n = 4) |>
  dummy_cols() |>
  select(!League)
```

### Create our workflow

```{r}
lasso_wf <- workflow() |>
  add_recipe(lasso_recipe) |>
  add_model(lasso_model)
```

### Fit our model using our workflow

Don't forget to use *training* data when fitting.

```{r}
lasso_fit <- lasso_wf |> fit(data = hitters_train)
```

### Review our coefficients

[`parsnip`](https://parsnip.tidymodels.org/) is a package in the `tidymodels` ecosystem.

```{r}
lasso_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  slice_head(n = 4)
```

Which features were shrunk to zero?

```{r}
#| echo: false

lasso_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(estimate == 0)
```

#### Large coefficients

The coefficients with the largest absolute values are the most important predictors of salary in this data set with this penalty:

```{r}
lasso_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  arrange(desc(abs(estimate))) |>
  slice_head(n = 5)
```

#### Interpreting coefficients

```{r}
#| echo: false

cruns_est <- lasso_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term == "CRuns") |>
  pull(estimate) * 1000
```

Since our features are normalized such that mean = 0 and standard deviation = 1, we interpret the coefficients like so:

For every 1 standard deviation above the mean value of career runs (`CRuns`), the average salary increases by `{r} scales::dollar(cruns_est)`.

### Evaluate our model

```{r}
lasso_fit |>
  predict(new_data = hitters_test) |>
  bind_cols(hitters_test) |>
  metrics(truth = Salary, estimate = .pred)
```

#### What are those metrics?

| **Metric** | **Name**                             | **What It Measures**                                                                                      |
|------------|--------------------------------------|------------------------------------------------------------------------------------------------------------|
| `rmse`     | [Root Mean Squared Error](https://en.wikipedia.org/wiki/Root_mean_square_deviation)          | The square root of the average of the squared differences between predicted and actual values.             |
| `rsq`      | [R Squared (Coefficient of determination)](https://en.wikipedia.org/wiki/Coefficient_of_determination) | The proportion of variance in the outcome that is explained by the model.                                  |
| `mae`      | [Mean Absolute Error](https://en.wikipedia.org/wiki/Mean_absolute_error)              | The average of the absolute differences between predicted and actual values.                               |

#### How do we use these metrics?

- We select a metric to optimize depending on the data and its use.
- Given an optimization metric, it is possible to automatically select the best lambda penalty for our model.
- We can also use these metrics to compare this model to other models.

#### When do we optimize for a particular metric?

- Use Root Mean Squared Error if large errors are particularly costly
  - e.g., predicting expenses, because large errors can lead to large financial losses
- Use $R^2$ if we want a measure of explained variance.
  - e.g., predicting the extent to which additional hours of studying help students pass exams
- Use Mean Absolute Error if each error is equally important
  - e.g., predicting drive time in Google Maps

## Automating lambda selection

In the previous example, we arbitrarily chose a lambda value of 0.5. Instead, we can `tune()` to automatically select the best lambda. This is a form of [hyperparameter tuning](https://en.wikipedia.org/wiki/Hyperparameter_optimization).

```{r}
lasso_tune_model <- linear_reg(penalty = tune(), mixture = 1) |>
  set_mode("regression") |>
  set_engine("glmnet")

lasso_tune_wf <- workflow() |>
  add_recipe(lasso_recipe) |>
  add_model(lasso_tune_model)
```

### Cross-validation

[Cross-validation](https://en.wikipedia.org/wiki/Cross-validation_(statistics)) systematically evaluates different lambda values and finds the one that minimizes prediction error. By splitting the training data into folds (`v = 10`), the model is trained on $n-1$ folds and tested on the remaining fold, repeating the process across all possible splits. This helps prevent overfitting and ensures that the selected lambda generalizes well to new data.

```{r}
hitters_folds <- vfold_cv(hitters_train, v = 10, strata = Salary)
```

### Tune over a grid of penalty values

`tune_grid()` fits the lasso model across the penalty values given in `grid`, using the cross-validation folds we just created:

```{r}
tuned_results <- tune_grid(
  lasso_tune_wf,
  resamples = hitters_folds,
  grid = 50 # number of penalty values to try
)
```

#### Visualize tuning process

```{r}
autoplot(tuned_results)
```

### Identify the best penalty value

We choose a metric to optimize for with `select_best()`:

```{r}
best_lasso <- select_best(tuned_results, metric = "rmse")

best_lasso
```

### Run workflow with the best penalty

```{r}
final_lasso_wf <- finalize_workflow(lasso_tune_wf, best_lasso)
```

### Fit the finalized workflow and evaluate on the test set

```{r}
final_lasso_fit <- final_lasso_wf |> fit(data = hitters_train)

final_lasso_fit |>
  predict(new_data = hitters_test) |>
  bind_cols(hitters_test) |>
  metrics(truth = Salary, estimate = .pred)
```

### Is that `rmse` any good?

It could be better. The root mean squared error is about half of the mean salary.

```{r}
mean_salary <- hitters_train |>
  summarize(mean_salary = mean(Salary)) |>
  pull()

rmse <- final_lasso_fit |>
  predict(new_data = hitters_test) |>
  bind_cols(hitters_test) |>
  metrics(truth = Salary, estimate = .pred) |>
  filter(.metric == "rmse") |>
  pull(.estimate)

rmse / mean_salary
```

## Elastic net

Let's see if we can do better with elastic net regularization.

### Create our elastic net model

You will notice that the only difference between the lasso and elastic net models is the `mixture` parameter, which we now `tune()` as well:

```{r}
enet_model <- linear_reg(penalty = tune(), mixture = tune()) |>
  set_mode("regression") |>
  set_engine("glmnet")
```

### Create our elastic net workflow

We can use the same recipe as before, so we will move straight to the workflow:

```{r}
enet_wf <- workflow() |>
  add_recipe(lasso_recipe) |> # same recipe as before
  add_model(enet_model)
```

### Create cross-validation folds

```{r}
enet_folds <- vfold_cv(hitters_train, v = 10, strata = Salary)
```

### Tune over a grid of penalty values

```{r}
tuned_enet_results <- tune_grid(
  enet_wf,
  resamples = enet_folds,
  grid = 50
)
```

### Identify the best penalty value

```{r}
best_enet <- select_best(tuned_enet_results, metric = "rmse")
```

### Run workflow with the best penalty

```{r}
final_enet_wf <- finalize_workflow(enet_wf, best_enet)
```

### Fit the finalized workflow and evaluate on the test set

```{r}
final_enet_fit <- final_enet_wf |> fit(data = hitters_train)

final_enet_fit |>
  predict(new_data = hitters_test) |>
  bind_cols(hitters_test) |>
  metrics(truth = Salary, estimate = .pred)
```

This performs slightly better than lasso, but we could do better with further tuning.

## Summary

Regularization controls model complexity by shrinking or eliminating variables.

  - **Ridge (`mixture = 0`)** shrinks coefficients toward zero but typically keeps them all in the model.
  - **Lasso (`mixture = 1`)** zeros out some coefficients.
  - **Elastic net (`0 < mixture < 1`)** combines the strengths of both lasso and ridge.
- **Tuning** the penalty (`lambda`) balances bias and variance.
- **Interpretation** of coefficients after normalization requires us to recall that each numeric predictor is on a standardized scale (mean = 0, standard deviation = 1).
- **Performance** is evaluated using error metrics like RMSE, $R^2$, and MAE.

## How could you use regularization for Project 2?

- Given the Taylor Swift lyrics and those of one other artist, predict whether an unlabeled song was written by Swift, *and* how the model "knows."
  - Lasso and ENET can tell you which words are most predictive of e.g., Swift authorship.
- Predict whether a song is from a particular year (continuous variable) or "era" (categorical variable) based on its lyrics.
- Predict Spotify streams using lyrics or other features from the corpus.
