---
title: "Corpora and Sampling"
subtitle: "DSST389: Advanced Data Science"
author: "Erik Fredner"
institute: "University of Richmond"
date: today
bibliography: /Users/erik/book/references.bib
csl: /Users/erik/code/styles/chicago-fullnote-bibliography.csl
execute:
  echo: true
  warning: false
  message: false
format:
  revealjs:
    logo: "images/by-sa.png"
    footer: "https://fredner.org"
    embed-resources: true
    scrollable: true
    toc: true
    toc-depth: 2
    slide-level: 4
    slide-number: true
    preview-links: auto
    mermaid:
      theme: neutral
editor_options:
  markdown:
    wrap: 72
---

## Review

### Poll Everywhere

1. Go to this URL:

<https://pollev.com/fredner>

2. Please log in to Poll Everywhere using your `@richmond.edu` email address.

### Notebook 3 review

#### Data

```{r}
library(tidyverse)
theme_set(theme_minimal())

funds <- read_csv("../data/chicago_spending.csv")

funds |> slice_sample(n = 5)
```

#### Parse number

```{r}
funds |>
  select(cost) |>
  slice_sample(n = 5) |>
  pull()

funds <- funds |>
  mutate(cost = parse_number(cost)) |>
  select(year, ward, cost, category)

funds |> slice_sample(n = 5)
```

#### Any blank values?

```{r}
funds |>
  filter(if_any(everything(), is.na))
```

#### Total expenditures

```{r}
funds |>
  group_by(category) |>
  summarize(total_cost = sum(cost)) |>
  mutate(prop_cost = total_cost / sum(total_cost)) |>
  arrange(desc(total_cost))
```

#### Expenditures stacked

```{r}
#| output-location: slide

funds |>
  group_by(year, category) |>
  summarize(total_cost = sum(cost)) |>
  ggplot(aes(x = as_factor(year), y = total_cost, fill = category)) +
  geom_col(position = "stack") +
  scale_fill_brewer(palette = "Set3")
```

#### A note about qualitative color palettes

```r
scale_fill_brewer(palette = "Set3")
```

This is a *qualitative* color palette, which is designed for categorical data and does not imply any inherent order or magnitude, unlike continuous continuous scales like `viridis`.

#### Expenditures stacked with continuous color palette (`viridis`)

:::{.callout-warning}
This color palette improperly implies a scale of values from low (dark) to high (light). Here, the factor order (and thus the color) is determined **alphabetically**, which is not meaningful.
:::

```{r}
#| output-location: slide
#| echo: false

funds |>
  group_by(year, category) |>
  summarize(total_cost = sum(cost)) |>
  ggplot(aes(x = as_factor(year), y = total_cost, fill = category)) +
  geom_col(position = "stack") +
  scale_fill_viridis_d()
```

#### Recommended color palettes

- For continuous variables (e.g., grade is between 0 and 100), the [`viridis` scale](https://ggplot2.tidyverse.org/reference/scale_viridis.html) is recommended
- For categorical variables (e.g., major is Biology, English, Philosophy, etc.), the colorblind-safe [`brewer` scales](https://ggplot2.tidyverse.org/reference/scale_brewer.html) are recommended (e.g., `"Dark2"`)
  - If you don't like `"Dark2"`, try `"Set2"` or `"Pastel2"`
- (I used `"Set3"` in the previous figure because `"Dark2"` only supports up to seven colors.)
  
#### `brewer` qualitative scales
  
```{r}
#| echo: false

library(ggplot2)
library(patchwork)

# Convert cyl and gear to factors
mtcars$cyl <- factor(mtcars$cyl)
mtcars$gear <- factor(mtcars$gear)

# Plot 1: Dark2
p_dark2 <- ggplot(mtcars, aes(x = cyl, fill = gear)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Dark2")

# Plot 2: Set2
p_set2 <- ggplot(mtcars, aes(x = cyl, fill = gear)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Set2")

# Plot 3: Pastel2
p_pastel2 <- ggplot(mtcars, aes(x = cyl, fill = gear)) +
  geom_bar() +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Pastel2")

# Combine all three plots in a 1x3 grid
p_dark2 | p_set2 | p_pastel2
```

#### Proportional expenditures area

```{r}
#| output-location: slide


funds |>
  group_by(year, category) |>
  summarize(total_cost = sum(cost)) |>
  ungroup() |>
  group_by(year) |>
  mutate(prop_cost = total_cost / sum(total_cost)) |>
  ggplot(aes(x = year, y = prop_cost, fill = category)) +
  geom_area(position = "stack") +
  scale_fill_brewer(palette = "Set3")
```

#### Proportional expenditures area simplified grouping

```{r}
#| output-location: slide

funds |>
  group_by(year, category) |>
  # .groups = "drop_last" drops `category` grouping, but keeps `year`:
  summarize(total_cost = sum(cost), .groups = "drop_last") |>
  mutate(prop_cost = total_cost / sum(total_cost)) |>
  ggplot(aes(x = year, y = prop_cost, fill = category)) +
  geom_area(position = "stack") +
  scale_fill_brewer(palette = "Set3")
```

#### Spending by ward

```{r}
funds |>
  group_by(ward) |>
  summarize(total_spend = sum(cost)) |>
  ungroup() |>
  filter(!is.na(total_spend)) |>
  summarize(
    mean_total_spend = mean(total_spend),
    sd_total_spend = sd(total_spend),
    cv_total_spend = sd_total_spend / mean_total_spend
  )
```

- `sd` abbreviates [standard deviation](https://en.wikipedia.org/wiki/Standard_deviation)
- `cv` abbreviates [coefficient of variation](https://en.wikipedia.org/wiki/Coefficient_of_variation)

#### Spending outliers

```{r}
ward_spending <- funds |>
  group_by(category, ward) |>
  summarize(total_spend = sum(cost), .groups = "drop")

ward_outliers <- ward_spending |>
  group_by(category) |>
  mutate(
    Q1 = quantile(total_spend, 0.25),
    Q3 = quantile(total_spend, 0.75),
    IQR = Q3 - Q1,
    upper_bound = Q3 + 1.5 * IQR
  ) |>
  filter(total_spend >= upper_bound) |>
  ungroup()

ward_outliers |>
  mutate(spend_over_upper = total_spend - upper_bound) |>
  arrange(desc(spend_over_upper)) |>
  slice_head(n = 1)
```

The Parks in Ward 1 include [*Cloud Gate*](https://en.wikipedia.org/wiki/Cloud_Gate).

#### Spending equity

We assume an equal split across all wards, and see how each ward’s actual spending deviates from that ideal.

```{r}
#| output-location: slide

total_spending <- funds |>
  summarize(total_cost = sum(cost)) |>
  pull(total_cost)

ward_spending <- funds |>
  group_by(ward) |>
  summarize(ward_total = sum(cost), ) |>
  ungroup()

n_wards <- funds |>
  summarize(max_ward = max(ward)) |>
  pull(max_ward)

expected_spending <- total_spending / n_wards

ward_spending <- ward_spending |>
  mutate(
    expected_spend = expected_spending,
    deviation = ward_total - expected_spend
  ) |>
  arrange(desc(deviation))

ward_spending |>
  ggplot(aes(x = fct_reorder(factor(ward), deviation), y = deviation)) +
  geom_col() +
  labs(
    title = "Deviation of Wards' Spending from Expected",
    subtitle = "Positive values indicate overspending, negative values indicate underspending",
    x = "Ward",
    y = "Spending Deviation (Actual - Expected)"
  ) +
  theme(axis.text.x = element_text(angle = 90))
```

#### Questions?

## Corpora and sampling

### Analyzing textual data

- If you want to study one text, just read it!
- Quantitative text analysis becomes interesting when you have:
  - a large enough amount of text that it would be difficult to read and recall all of the relevant details
  - a large enough amount of text that it would be impossible to read and recall all of the relevant details
- Neither of these mean that you need to have an exceptionally large amount of text.
  - e.g., Human readers identifying parts of speech

### Corpora

- The set of documents being studied is often called the **corpus** (plural: corpora)
- What's "in" and what's "out" of the corpus?

#### Example: Taylor Swift corpora

- Corpora containing lyrics to...
  - "Cruel Summer" (one song, but with multiple versions)
  - *Red (Taylor's Version)* (one album)
  - all of the *Taylor's Version* albums (four albums)
  - all of Swift's recorded songs (Studio only? Live? YouTube?)
  - Swift's songs, compared with other pop songs (???)

### Problems with corpus construction

1. Completeness
2. Reproducibility
3. Representativeness

### Corpora as an example of a general problem

These issues with corpus construction are a specific example of a general problem in data science, which pertains to the representativeness of the *sample* we have in relation to the *population* we want to measure.

#### Bigger is not necessarily better

- [The *Literary Digest* poll of 1936](https://doi.org/10.1017/S014555320001035X) famously predicted that Alf Landon would defeat Franklin D. Roosevelt in the US presidential election
  - *Literary Digest* polled over 10 million people, but they oversampled Landon supporters (sample bias)
- George Gallup (later of Gallup polling) conducted biweekly polls of just 2,000 people and accurately predicted a Roosevelt victory
  - Quality of the sample beats quantity

#### Selection bias

- *Literary Digest* subscribers were different from the population of US voters in a non-random way
- The same is true for contemporary rating systems, like Yelp and Google reviews (self-selection bias)

### Important distributions

Once we have some samples, we will take some sample statistics. Here are two important ideal types for these distributions.

#### Uniform distribution

This assumes that every number between two values is equally likely to be chosen. [**Bingo**](https://en.wikipedia.org/wiki/Bingo_(American_version)) is a good real-world example.

```{r}
set.seed(1)

bingo_calls <- tibble(
  call_order = 1:75,
  ball = sample(1:75, size = 75, replace = FALSE)
)

bingo_calls |> slice_head(n = 5)
```

#### Bingo and uniformity

Let's see when Ball 1 would be called over 100,000 games of bingo:

```{r}
#| echo: false

set.seed(1)
n_games <- 100000
n_balls <- 75

bingo_sim <- tibble(game = 1:n_games) %>%
  mutate(call_order_ball1 = map_int(game, ~ {
    call_sequence <- sample(1:n_balls, size = n_balls, replace = FALSE)
    which(call_sequence == 1)
  }))
```

```{r}
bingo_sim |> slice_sample(n = 5)
```

```{r}
#| output-location: slide

bingo_sim |>
  ggplot(aes(x = call_order_ball1)) +
  geom_histogram(binwidth = 1, color = "white", fill = "#009E73") +
  scale_x_continuous(
    breaks = 1:n_balls,
    guide = guide_axis(check.overlap = TRUE)
  )
```

#### Normality

Unlike the uniform distribution, the [normal (or Gaussian) distribution](https://en.wikipedia.org/wiki/Normal_distribution) is unequal. **Human height** is a good real-world example.

```{r}
set.seed(1)

n <- 1000
mean_height <- 165 # in centimeters
sd_height <- 7 # in centimeters

women_heights <- tibble(
  student_id = 1:n,
  height = rnorm(n, mean = mean_height, sd = sd_height)
) |>
  mutate(
    total_inches = height / 2.54,
    feet = floor(total_inches / 12),
    inches = total_inches - feet * 12
  ) |>
  select(-total_inches)

women_heights |> slice_sample(n = 5)
```

#### The bell curve

```{r}
#| echo: false

women_heights |>
  ggplot(aes(x = height)) +
  geom_histogram(
    aes(y = ..density..),
    binwidth = 1,
    fill = "white",
    color = "black"
  ) +
  geom_density(fill = "#009E73", alpha = 0.2) +
  stat_function(
    fun = dnorm, args = list(mean = mean_height, sd = sd_height),
    color = "#56B4E9", size = 1
  ) +
  labs(
    title = "Simulated heights of female American college students",
    subtitle = "Assuming a normal distribution (mean = 165 cm, sd = 7 cm)",
    x = "Height (cm)",
    y = "Density"
  )
```

#### A challenge with word distributions in texts

- Word distributions are not uniform because not all words are equally likely.
- Not all word distributions are normal because many words are "bursty."
  - The word "Dracula" shows up a lot in [*Dracula*](https://en.wikipedia.org/wiki/Dracula).
  - But that large number of uses does not mean that we should expect to see a few instances of "Dracula" in many novels from 1897.

### Practice
